# ============================================
---
title: "LA Urban Heat & Carbon Risk Explorer"
subtitle: "Multi-layer Environmental Burden Analysis: NDVI, Pollution, and Electricity"
author: "Qingrong (Serena) Li"
date: "May 2026"
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    theme: darkly
---
# Setup
# ============================================
```{r}
library(duckdbfs)
library(dplyr)
library(rstac)
library(gdalcubes)
library(sf)
library(mapgl)
library(ellmer)
library(readxl)
library(ggplot2)
gdalcubes_options(parallel = TRUE)
```
# ============================================
# 1. Load HOLC Redlining Data
# ============================================
```{r}
holc_la <- open_dataset(
  "https://data.source.coop/cboettig/mappinginequality/mappinginequality.parquet"
) |>
  filter(city == "Los Angeles") |>
  to_sf(crs = 4326)

bbox <- st_bbox(holc_la)

holc_clean <- holc_la |>
  select(area_id, grade, category, geom) |>
  st_transform("EPSG:4326")
```
# ============================================
# 2. Query Sentinel-2 via STAC
# ============================================
```{r}
items <- stac("https://earth-search.aws.element84.com/v1/") |>
  stac_search(
    collections = "sentinel-2-c1-l2a",
    bbox        = as.numeric(bbox),
    datetime    = "2023-06-01T00:00:00Z/2023-08-31T23:59:59Z",
    limit       = 200
  ) |>
  ext_query("eo:cloud_cover" < 20) |>
  post_request() |>
  items_fetch()
cat(length(items$features), "scenes found\n")
```
# ============================================
# 3. Compute NDVI
# ============================================
```{r}
col <- stac_image_collection(
  items$features,
  asset_names = c("red", "nir", "scl")
)
v <- cube_view(
  srs        = "EPSG:4326",
  extent     = list(
    t0 = "2023-06-01", t1 = "2023-08-31",
    left   = bbox["xmin"], right  = bbox["xmax"],
    bottom = bbox["ymin"], top    = bbox["ymax"]
  ),
  dx = 0.0002, dy = 0.0002,
  dt = "P3M",
  aggregation = "median",
  resampling  = "bilinear"
)
S2.mask <- image_mask("scl", values = c(3, 8, 9))
ndvi <- raster_cube(col, v, mask = S2.mask) |>
  select_bands(c("red", "nir")) |>
  apply_pixel("(nir-red)/(nir+red)", "NDVI") |>
  reduce_time("mean(NDVI)")
```
# ============================================
# 4. Zonal Statistics per HOLC Area
# ============================================
```{r}
zonal <- ndvi |> extract_geom(holc_clean, FUN = mean, reduce_time = TRUE)
holc_clean$NDVI <- zonal$NDVI
holc_clean$carbon_score <- 1 - holc_clean$NDVI

holc_clean |>
  st_drop_geometry() |>
  group_by(grade) |>
  summarise(
    mean_NDVI  = round(mean(NDVI, na.rm = TRUE), 3),
    mean_score = round(mean(carbon_score, na.rm = TRUE), 3),
    n = n()
  )
```
# ============================================
# 5. Electricity & Carbon Emissions
# ============================================
```{r}
elec_raw <- read_excel("data/ca_electricity_county.xlsx")

elec_la <- elec_raw |>
  filter(COUNTY_NAME == "LOS ANGELES", YEAR == 2023) |>
  group_by(SECTOR) |>
  summarise(total_GWh = sum(GWH, na.rm = TRUE)) |>
  arrange(desc(total_GWh)) |>
  mutate(
    total_MWh          = total_GWh * 1000,
    carbon_emissions_MT = total_MWh * 0.091
  )

ggplot(elec_la, aes(x = reorder(SECTOR, carbon_emissions_MT),
                    y = carbon_emissions_MT / 1e6,
                    fill = SECTOR)) +
  geom_col() +
  coord_flip() +
  labs(
    title    = "LA County Carbon Emissions by Sector (2023)",
    subtitle = "Electricity × CA grid carbon intensity (0.091 MT CO2/MWh)",
    x = NULL,
    y = "Carbon Emissions (Million MT CO2)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::comma)
```
# ============================================
# 6. CalEnviroScreen - Pollution Score
# ============================================
```{r}
calenviro_url <- paste0(
  "https://services1.arcgis.com/PCHfdHz4GlDNAhBb/arcgis/rest/services/",
  "CalEnviroScreen_4_0_Results_/FeatureServer/0/query?",
  "where=1%3D1&outFields=tract%2CCIscore%2CCIscoreP%2CTractTXT",
  "&returnGeometry=true&f=geojson"
)

calenviro_all <- read_sf(calenviro_url)

calenviro_la <- calenviro_all |>
  filter(substr(TractTXT, 1, 4) == "6037") |>
  st_transform(4326)
```
# ============================================
# 7. Spatial Join: HOLC + CalEnviroScreen
# ============================================
```{r}

sf_use_s2(FALSE)

calenviro_la_valid <- st_make_valid(calenviro_la)
holc_clean_valid <- st_make_valid(holc_clean)

holc_with_pollution <- st_join(
  holc_clean_valid,
  calenviro_la_valid |> select(CIscore, TractTXT),
  join = st_intersects
)

summary(holc_with_pollution$CIscore)
```
# ============================================
# 8. Three-Layer Map
# Color= NDVI，Height = CIscore
# ============================================
```{r}
holc_with_pollution <- holc_with_pollution |>
  mutate(
    CIscore_clean = ifelse(is.na(CIscore), 0, CIscore),
    # 指数放大：每差1分，高度差2倍效果
    height = exp(CIscore_clean / 15) * 50
  )

summary(holc_with_pollution$height)

maplibre(
  style  = carto_style("dark-matter"),
  center = c(mean(bbox[c("xmin","xmax")]), mean(bbox[c("ymin","ymax")])),
  zoom   = 10,
  pitch  = 60,
  projection = "mercator"
) |>
  add_fill_extrusion_layer(
    id     = "combined",
    source = holc_with_pollution,
    fill_extrusion_color = step_expr(
      column = "NDVI",
      base   = "#d73027",
      values = c(0.1, 0.15, 0.2, 0.25),
      stops  = c("#fc8d59", "#ffffbf", "#91cf60", "#1a9850")
    ),
    fill_extrusion_height = step_expr(
      column = "height",
      base   = 10,
      values = c(100, 300, 600, 1000),
      stops  = c(100, 300, 600, 1000)
    ),
    fill_extrusion_opacity = 0.9,
    tooltip = "CIscore"  
  ) |>
  add_legend(
    "Color = NDVI | Height = Pollution Burden",
    values = c("Low NDVI (Red)", "High NDVI (Green)"),
    colors = c("#d73027", "#1a9850")
  )
```

# ============================================
# 9. AI Analysis
# ============================================
```{r}
Sys.setenv(NRP_API_KEY = "KWEsqVLGHqI3nguEHrWdEay8TVV4CoWz")
chat <- chat_openai_compatible(
  base_url    = "https://ellm.nrp-nautilus.io/v1",
  credentials = function() Sys.getenv("NRP_API_KEY"),
  model       = "qwen3-small"
)

summary_stats <- holc_with_pollution |>
  st_drop_geometry() |>
  group_by(grade) |>
  summarise(
    mean_NDVI    = round(mean(NDVI, na.rm = TRUE), 3),
    mean_CIscore = round(mean(CIscore, na.rm = TRUE), 1)
  ) |>
  as.data.frame()

chat$chat(paste(
  "In Los Angeles, NDVI and pollution burden (CalEnviroScreen score) by HOLC grade:",
  paste(capture.output(print(summary_stats)), collapse = " "),
  "Carbon emissions by sector (2023, million MT CO2):",
  paste(capture.output(
    print(elec_la |>
      select(SECTOR, carbon_emissions_MT) |>
      mutate(carbon_emissions_MT = round(carbon_emissions_MT / 1e6, 2)))
  ), collapse = " "),
  "What does this combined data tell us about environmental justice",
  "and where should green infrastructure and carbon reduction be prioritized?"
))
```
```{r}
st_write(holc_with_pollution, 
         "data/holc_pollution.geojson", 
         delete_dsn = TRUE)
```
